---
title: "Mediation Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

Causal inference often requires investigating *how* an effect occurs. Does $X$ affect $Y$ directly, or does it work through a mediator $M$?

The `because` package provides a fully automated Bayesian mediation analysis tool, `because_mediation()`, which decomposes the Total Effect of an exposure on an outcome into: 1. **Direct Effect**: The effect of $X \to Y$ not mediated by other variables in the graph. 2. **Indirect Effect(s)**: The effect propagated through intermediate variables ($X \to M \to Y$).

This is calculated by multiplying the posterior distributions of coefficients along each path, preserving full uncertainty quantification.

# Example: Ecological Mediation (Elevation Gradient)

In this example, we investigate how **Elevation** affects **Plant Abundance**. We hypothesize that Elevation acts through a causal chain involving Temperature and Soil Moisture:

1.  **Elevation** determines **Temperature** (higher elevation $\to$ lower temperature).
2.  **Temperature** influences **Soil Moisture** (lower temperature $\to$ lower evaporation $\to$ higher moisture).
3.  **Abundance** is driven by **Moisture**, **Temperature**, and potentially a direct effect of **Elevation** (e.g., due to UV radiation or partial pressure of gases).

## 1. Simulate Data

We simulate $N=200$ plots along an elevation gradient.

```{r setup}
library(because)

set.seed(42)
N <- 200

# 1. Elevation (Exogenous variable)
# Ranges roughly from 500m to 1500m
Elevation <- rnorm(N, mean = 1000, sd = 200)

# 2. Temperature (Mediator 1)
# Decreases with Elevation (Lapse rate approx effect)
# Coef: -0.01 implies 100m climb -> -1 degree C
Temp <- 25 - 0.01 * Elevation + rnorm(N, sd = 2)

# 3. Moisture (Mediator 2)
# Driven by Temperature. Cooler -> Moister.
# We model a negative relationship with Temp.
Moisture <- 20 - 2 * Temp + rnorm(N, sd = 5)

# 4. Plant Abundance (Outcome)
# - Positive effect of Moisture (+0.5)
# - Positive effect of Temperature (+1.5)
# - Direct negative effect of Elevation (-0.005) due to harsh conditions
Abundance <- 10 + 0.5 * Moisture + 1.5 * Temp - 0.005 * Elevation + rnorm(N, sd = 10)

eco_data <- data.frame(Elevation, Temp, Moisture, Abundance)
head(eco_data)
```

## 2. Standardize Data

**Important:** For mediation analysis, it is highly recommended to **standardize** your continuous variables (mean = 0, sd = 1) before fitting. 

Standardization ensures that:
1.  **Coefficients are comparable**: All effects are expressed in standard deviation units (standardized effects).
2.  **Scale Invariance**: The calculation of Indirect Effects (product of coefficients) is more interpretable relative to the Total Effect.
3.  **Convergence**: MCMC sampling often behaves better with standardized scales.

```{r standardize}
# Standardize all variables
eco_data_std <- scale(eco_data)
head(eco_data_std)
```

## 3. Fit the Structural Equation Model

We define the structural equations reflecting our causal DAG. Notice the chain: `Elevation -> Temp -> Moisture -> Abundance`.

```{r fit_model}
# Define the structural equations
eco_eqs <- list(
  Temp ~ Elevation,
  Moisture ~ Temp,
  Abundance ~ Moisture + Temp + Elevation
)

# Fit the model
# We use a short chain for demonstration purposes. Use more iterations for real analysis.
fit <- because(
  equations = eco_eqs,
  data = eco_data_std,
  n.iter = 2000
)
summary(fit)
```

We can also plot the fitted causal model with its standardised paths:
```{r plot_dag}
plot_dag(fit)
```

## 4. Perform Mediation Analysis

We want to understand the Total Effect of **Elevation** on **Abundance**, and decompose it into its direct and indirect components.

```{r mediation}
# Run Mediation Analysis for Elevation -> Abundance
med_results <- because_mediation(fit, exposure = "Elevation", outcome = "Abundance")
```

### Inspect the Summary

```{r summary}
med_results$summary
```

**Interpretation (Standardized Units):**
*   **Total Effect**: The net impact of Elevation on Abundance in SD units.
*   **Direct Effect**: The path `Elevation -> Abundance` (independent of mediators).
*   **Total Indirect Effect**: The sum of all mediated paths. In our simulation, Elevation lowers Temp, which raises Moisture, which increases Abundance.

### Inspect Individual Paths

The `because_mediation` function automatically traces all valid paths from exposure to outcome in the DAG.

```{r paths}
med_results$paths
```

We expect to see three distinct paths:

1.  **Direct**: `Elevation -> Abundance`
2.  **Short Indirect**: `Elevation -> Temp -> Abundance`
    *   Elevation lowers Temp (negative correlation); Temp increases Abundance (positive correlation).
    *   The product of these effects is **Negative**.
3.  **Long Indirect (Chain)**: `Elevation -> Temp -> Moisture -> Abundance`
    *   Elevation lowers Temp (negative).
    *   Lower Temp raises Moisture (negative relationship $\to$ positive change in moisture).
    *   Higher Moisture raises Abundance (positive).
    *   The chain involves two negative links and one positive link, resulting in a **Positive** indirect effect.

The function handles this decomposition automatically, providing credibility intervals for each specific mechanism.

## Technical Note

The function calculates the indirect effect as the product of coefficients along the path. For example, for the long chain: 
$$ \text{Indirect}_{\text{chain}} = \beta_{Elev \to Temp} \times \beta_{Temp \to Moist} \times \beta_{Moist \to Abund} $$ 
This approach is standard for linear models. For non-linear models, this represents an approximation of the average causal effect.
