---
title: "Multilevel Causal Models"
author: "Achaz von Hardenberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multilevel Causal Models}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

Many ecological and evolutionary datasets are inherently multilevel: observations of individuals are nested within years, years within sites, or individuals within populations. In such settings, variables measured at different levels of the hierarchy have fundamentally different sample sizes and thus different statistical power. A climate covariate averaged over 10 years provides only 10 degrees of freedom for testing its effects, regardless of how many individual observations exist within those years.

The `because` package explicitly handles this structure. When data are supplied as a **named list of data frames**, one per hierarchical level, the engine:

1.  Correctly maps each variable to its natural level.
 
2.  Uses the **appropriate degrees of freedom** for each d-separation test: year-level tests use $N_{years}$, individual-level tests use $N_{individuals}$, and cross-level tests are fitted at the lower level using whatever random effects the researcher specified in the structural equations.
 
3.  Skips tests that are trivially satisfied by the design of the hierarchy (e.g., orthogonal branches that share no observations).

## Example: Climate, Resources, and Reproductive Success

Consider a long-term monitoring study of a wild mammal population. Over 10 years, ecologists record:

-   **Year-level** variables (1 row per year): mean winter temperature (`Temp`) and annual primary productivity (`NDVI`).
-   **Individual-level** variables (multiple rows per individual per year): body mass (`Mass`) and number of offspring (`Offspring`).

The causal hypothesis is:

-   Winter temperature controls primary productivity (`Temp → NDVI`).
-   Productivity and temperature together drive individual body mass (`NDVI → Mass`, `Temp → Mass`).
-   Body mass determines reproductive success (`Mass → Offspring`).

This is a cross-level causal chain: climate acts at the year level, but its ultimate effect on offspring is mediated through individual body condition.

```{r dag}
library(because)

dag <- list(
  NDVI       ~ Temp,
  Mass       ~ NDVI + Temp,
  Offspring  ~ Mass
)

plot_dag(dag)
```

## Data Preparation

To use the hierarchical modeling features, your data must be structured as a **named list of data frames**, where each list element corresponds to a level in your hierarchy.

### The "One Home" Rule

Every variable in your model (predictors and responses) must live in **exactly one** data frame (the one corresponding to the level where it was measured). 

*   **Year-level variables** (like `Temp`) should only be in the `year` data frame.
*   **Individual-level variables** (like `Mass`) should only be in the `individual` data frame.

Do **not** repeat year-level variables in the individual-level data frame. `because` will automatically handle the cross-level mapping using your `link_vars`.

### Auto-detection vs. Manual Levels

If you omit the `levels` argument, `because` will try to auto-detect which variable belongs to which level. This works well for simple models, but providing the `levels` list manually is recommended because:
1.  It resolves any ambiguity if multiple data frames share column names.
2.  It allows you to explicitly **exclude** nuisance variables that might be in your data frames but are not part of the causal model.

## Simulate Hierarchical Example Data

```{r simulate}
set.seed(42)

n_years <- 10
n_ind   <- 10   # 10 individuals followed over time

# --- Year-level data ---
year_df <- data.frame(
  Year = 1:n_years,
  Temp = rnorm(n_years, mean = 0, sd = 1),
  NDVI = NA_real_
)
year_df$NDVI <- 0.7 * year_df$Temp + rnorm(n_years, 0, 0.5)

# --- Individual-level (Repeated Measures) data ---
# We use expand.grid to ensure every individual is measured every year
ind_df <- expand.grid(
  ID   = 1:n_ind,
  Year = 1:n_years
)

# For simulation, each individual has a unique "personality" (random effect)
# that stays constant across years.
ind_personality <- rnorm(n_ind, 0, 0.5)
names(ind_personality) <- 1:n_ind

# Look up year-level and individual-level components
temp_lookup <- year_df$Temp[ind_df$Year]
ndvi_lookup <- year_df$NDVI[ind_df$Year]
pers_lookup <- ind_personality[as.character(ind_df$ID)]

# Mass depends on year-level NDVI/Temp AND the individual's personality
ind_df$Mass <- (
  1.2 * ndvi_lookup + 
  0.4 * temp_lookup + 
  pers_lookup +        # <-- Individual-level consistency
  rnorm(nrow(ind_df), 0, 0.4)
)

# Offspring depends on Mass
ind_df$Offspring <- rpois(nrow(ind_df), lambda = exp(0.5 + 0.6 * ind_df$Mass))

# Combine into a hierarchical list
data_list <- list(
  year       = year_df,
  individual = ind_df
)
```

## Supplying Multilevel Data to `because`

Pass the data as a **named list of data frames**. The names define the level labels that `because` uses to assign variables.

```{r data_list, eval = FALSE}
data_list <- list(
  year       = year_df,
  individual = ind_df
)
```

We specify the structural equations as usual:

```{r equations, eval = FALSE}
equations <- list(
  NDVI      ~ Temp,
  Mass      ~ NDVI + Temp,
  Offspring ~ Mass
)
```

and finally we can fit the model with `because()`, specifying the `hierarchy` and the  variable linking the two dataset (ie present in both datasets) with the `link_vars` argument. 

```{r fit, eval = FALSE}
fit <- because(
  equations  = equations,
  data       = data_list,
  family     = c(Offspring = "poisson"),
  hierarchy  = "year > individual",
  link_vars  = c(year = "Year"),
  n.iter     = 15000,
  n.burnin   = 3000
)

summary(fit)
```

## Understanding the Hierarchical Data Structure

### The `hierarchy` argument

`hierarchy = "year > individual"` tells `because` that the **year**
level is higher (coarser) and the **individual** level is nested within
it. The `>` operator means "is a parent of".

### The `link_vars` argument

`link_vars = c(year = "Year")` specifies that the column `Year` in the
**individual-level** data frame is the index that maps each individual
record back to its corresponding row in the **year-level** data frame.
This is analogous to a foreign key in a relational database.

### Automatic level assignment

`because` inspects which data frame each variable belongs to:

| Variable | Level | N (df for d-sep) |
|:---|:---|:---|
| `Temp` | year | 10 |
| `NDVI` | year | 10 |
| `Mass` | individual | 300 |
| `Offspring` | individual | 300 |


## Random Effects in `because`

`because` uses the same `(1|grouping_variable)` notation as `lme4`'s `lmer()` / `glmer()` to specify **random intercepts**. This will be
immediately familiar if you have used mixed-effects models in R.

A random intercept `(1|Group)` tells `because` that observations sharing the same value of `Group` are not independent — each group gets its own intercept drawn from a common Normal distribution:

$$u_g \sim \text{Normal}(0, \sigma^2_u), \quad g = 1, \dots, G$$

with the precision $\tau_u = 1/\sigma^2_u$ estimated from the data.

> [!NOTE]
> **Random slopes** (e.g. `(X|Group)`) are **not yet implemented**.
> Only random intercepts `(1|Group)` are currently supported. The syntax
> deliberately mirrors `lme4` so that random slopes can be added as a
> natural extension in a future release without any change to the user interface.

### Syntax option A: inline in the equation

Place `(1|Group)` directly inside any structural equation, just as you would in `lmer()`. You can specify different grouping structures for different responses:
For example, if you believe there is year-level variation in `Mass` that is not fully explained by `Temp` and `NDVI`, you can specify:

```{r re_per_equation, eval = FALSE}
equations_re <- list(
  NDVI      ~ Temp,  
  Mass ~ NDVI + Temp + (1|Year),    # Year random intercept for Mass
  Offspring ~ Mass  + (1|Year) + (1|ID)  # Year AND individual random intercepts
)
```

The random terms are **equation-specific**: here only `Offspring` gets the individual random intercept `(1|ID)`, while both `Mass` and `Offspring` get the year random intercept `(1|Year)`.

You would then run this model like this: 

```{r fit_inline_re, eval = FALSE}
 fit_re <- because(
  equations  = equations_re,
  data       = data_list,
  family     = c(Offspring = "poisson"),
  hierarchy  = "year > individual",
  link_vars  = c(year = "Year"),
  n.iter     = 15000,
  n.burnin   = 3000
)

summary(fit_re)
```

### Syntax option B: global `random` argument (convenient when all equations share the same grouping)**

We could specify the same model as above using the `random` argument to apply the same random structure to all equations (in this case the year random intercept applies to both equations, while the individual random intercept applies only to `Offspring` because `Mass` does not have `(1|ID)` in its equation): 

```{r re_per_equation, eval = FALSE}

equations_re2 <- list(
  Mass ~ NDVI + Temp,    # Year random intercept for Mass
  Offspring ~ Mass  + (1|ID)  # Year AND individual random intercepts
)

fit_re2 <- because(
  equations  = equations_re2,
  data       = data_list,
  family     = c(Offspring = "poisson"),
  hierarchy  = "year > individual",
  link_vars  = c(year = "Year"),
  random     = ~(1|Year), # applied to ALL equations
  dsep       = TRUE,
  n.iter     = 15000,
  n.burnin   = 3000
)
```

Both syntaxes are internally parsed into the same JAGS/NIMBLE model code. The inline
version gives you equation-level control; the global version is a shorthand when the grouping structure is uniform.

> [!NOTE]
> Random intercepts specified by the user — whether inline or global —
> are propagated into the d-separation sub-tests for all equations they
> apply to. This ensures the independence tests reflect exactly the same
> random structure as the structural model, maintaining internal
> consistency.

### When to include a random effect

| Situation | Include `(1|Group)`? |
|:---|:---|
| Observations share a year-level environment and year predictors **do not** fully capture year variation | **Yes** |
| Year-level predictors (e.g. Temp, NDVI) fully explain year variation | Not necessary |
| Repeated measures on the same individual across time | **Yes** — `(1|ID)` |
| Balanced, fully crossed design with all levels as fixed effects | Not needed |




## D-Separation with Correct Degrees of Freedom

The causal model implies the following basis set of conditional
independence claims (the minimal set of claims whose failure would
falsify the model):

| Test | Level | Conditioning set |
|:---|:---|:---|
| `NDVI _\|\|_ Mass \| {Temp}` | **year** (N=10) | Temp |
| `NDVI _\|\|_ Offspring \| {Mass, Temp}` | **individual** (N=300) | Mass, Temp |
| `Temp _\|\|_ Offspring \| {Mass, NDVI}` | **individual** (N=300) | Mass, NDVI |

> [!IMPORTANT]
> The first test — `NDVI _||_ Mass | {Temp}` — tests a year-level
> independence claim. Even though 300 individual observations exist,
> `because` correctly uses **N=10** degrees of freedom for this test,
> because both NDVI and Temp (the conditioning set) are year-level
> variables. Using individual-level N here would be pseudoreplication
> and would dramatically inflate statistical power.

### Within-level tests

For tests where **both** the response and the focal predictor are at the
**same level**, `because` fits the sub-model directly on that level's
data, naturally using the correct sample size.

### Cross-level tests

For tests where the response and focal predictor are at **different
levels** (e.g., testing whether a year-level predictor is independent of
an individual-level response, given a conditioning set), `because` fits
the sub-model at the **lower (individual) level** using the individual-level
data.

Crucially, the d-sep tests use **exactly the same random effect
structure as the structural equations**. If the researcher did not
specify a year-level random effect in their structural model, neither
will the d-sep tests. This is by design and scientifically coherent:

- The structural equation `Offspring ~ Mass` (without `(1|Year)`) embodies
  the scientific assumption that the year-level predictors already in the
  model — `Temp` and `NDVI` — fully account for year-level variation in
  the response. No residual year-level heterogeneity needs to be modelled.
- A d-sep test like `Offspring _||_ NDVI | {Mass, Temp}` conditions on `Temp`,
  a year-level variable. Conditioning on `Temp` statistically accounts for
  the year structure *within that test*, making an additional `(1|Year)` term
  redundant given the researcher's modelling assumptions.
- Automatically inserting year random effects into the d-sep tests when
  the researcher deliberately omitted them from the structural model would
  create an **internal inconsistency** — the test would be evaluated under
  stricter assumptions than the model itself.

> [!IMPORTANT]
> If you believe there are **unmeasured year-level confounders** beyond
> `Temp` and `NDVI`, you should include a year random effect explicitly in
> your structural equations (e.g., `random = ~(1|Year)` in `because()`).
> This will then also be correctly propagated into all relevant d-sep tests.

### Orthogonal branches (automatically skipped)

If the hierarchy contains parallel branches (e.g., `site > obs` and
`species > obs`) and the variables in a test come from different
branches with no conditioning variable bridging them, `because` will
skip the test.

This skip is a consequence of the **hierarchical specification**: the
d-separation engine is optimized for linear nesting chains. Tests
between orthogonal branches (which would require crossed random effects)
are currently treated as being outside the structural scope defined by
the linear hierarchy. This ensures the engine remains robust and avoids
attempting cross-level regressions where no shared hierarchical index
exists beyond the observation level.

## Visualising the Results

```{r plot_dag}
# DAG with estimated path coefficients
plot_dag(fit)
```

```{r plot_coef}
# Caterpillar plot of all path coefficients
plot_coef(fit)
```

```{r plot_dsep}
# D-separation results with 95% credible intervals
plot_dsep(fit)
```

## Practical Guidelines

### When to use multilevel data in `because`

-   Your predictors are measured at a **coarser temporal or spatial
    grain** than your response (e.g., annual climate → daily behaviour).
-   You have **repeated measures** on the same individuals across
    multiple occasions (years, seasons).
-   You want to partition effects into **between-group** (year-level)
    and **within-group** (individual-level) components.

### Minimal setup

At minimum, you need:

1.  A **named list** of data frames, one per level.
2.  A `hierarchy` string (e.g., `"year > individual"`).
3.  A `link_vars` named vector mapping the linking column to its parent
    level (e.g., `c(year = "Year")`).

If `levels` is omitted, `because` will automatically assign each
variable to the data frame it is found in.

### Variable overlap between levels

If the same column name appears in more than one data frame (e.g., a
site-level mean temperature and an individual-level body temperature
both named `Temp`), use distinct column names to avoid ambiguity.

## References

Shipley, B. (2016). *Cause and Correlation in Biology: A User's Guide
to Path Analysis, Structural Equations and Causal Inference with R* (2nd
ed.). Cambridge University Press.

Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and
Multilevel/Hierarchical Models*. Cambridge University Press.

Nakagawa, S., & Schielzeth, H. (2013). A general and simple method for
obtaining R² from generalized linear mixed-effects models. *Methods in
Ecology and Evolution*, 4(2), 133–142.
