---
title: "Creating Custom Distribution Families"
author: "Achaz von Hardenberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Custom Distribution Families}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

`because` natively supports the most common distribution families
(Gaussian, Poisson, Binomial, Negative Binomial, Ordinal, etc.). For
specialised data types, the `because_family()` function lets you define
any JAGS-compatible distribution as a drop-in replacement with no need
to modify the core package.

**When do you need a custom family?**

- **Robust regression**: Student-*t* for data with heavy tails or
  outliers.
- **Bounded proportions**: Beta distribution for continuous proportions
  in (0, 1) (e.g., percentage cover, survival rates).
- **Truncated distributions**: Modelling quantities that are physically
  bounded (e.g., speeds that cannot be negative, body temperatures
  within a viable range).
- **Specialised ecological distributions**: Log-normal, Gamma, or
  compound distributions not yet natively supported.

---

## The `because_family()` Constructor

```{r family_signature}
because_family(
  name,           # A short name (e.g., "student_t", "beta_prop")
  jags_likelihood,# JAGS likelihood code with placeholders
  link    = "identity",  # "identity", "log", or "logit"
  extra_priors = NULL,   # Additional JAGS prior statements
  description  = NULL    # Optional human-readable description
)
```

### Template Placeholders

The `jags_likelihood` string uses four placeholders that `because`
replaces at model-build time:

| Placeholder | Replaced with |
|:---|:---|
| `{response}` | The response variable name (e.g., `Cover`) |
| `{mu}` | The linear predictor mean vector (e.g., `mu_Cover`) |
| `{tau}` | The residual precision parameter (e.g., `tau_res_Cover`) |
| `{i}` | The loop index (`i`) |

`extra_priors` uses the same placeholders and is appended outside the
likelihood loop — use it for additional parameters like degrees of
freedom or dispersion.

---

## Example 1: Student-*t* Family (Robust Regression)

The Student-*t* distribution with $\nu$ degrees of freedom has heavier
tails than a Gaussian, making it robust to outliers. JAGS provides it
as `dt(mu, tau, nu)`.

```{r student_t_create}
library(because)

student_t <- because_family(
  name = "student_t",

  # JAGS dt(mu, tau, nu): location, precision, degrees of freedom
  jags_likelihood = "{response}[{i}] ~ dt({mu}[{i}], {tau}, df_{response})",

  link = "identity",

  # Prior on degrees of freedom: wide uniform between 2 (very heavy) and 100 (near-Gaussian)
  extra_priors = c("df_{response} ~ dunif(2, 100)"),

  description = "Student-t for robust regression with estimated degrees of freedom"
)
```

`because_family()` returns a **constructor function** (`student_t`).
Call it with no arguments to create the family object to pass to
`because()`:

```{r student_t_use}
set.seed(42)
N <- 80

X <- rnorm(N)

# Heavy-tailed response: many mild values plus a few large outliers
Y <- 1.5 * X + rt(N, df = 3)

robust_data <- data.frame(Y = Y, X = X)

# Gaussian model (potentially biased by outliers)
fit_gaussian <- because(
  equations = list(Y ~ X),
  data      = robust_data,
  quiet     = TRUE
)

# Student-t model (robust to outliers)
fit_robust <- because(
  equations = list(Y ~ X),
  data      = robust_data,
  family    = c(Y = student_t()),   # Pass the constructor result
  quiet     = TRUE
)

summary(fit_robust)
# df_Y: estimated degrees of freedom — small values (2–5) indicate heavy tails
```

### Comparing Fit

```{r student_t_compare}
fit_gaussian_waic <- because(
  equations = list(Y ~ X), data = robust_data, WAIC = TRUE, quiet = TRUE
)
fit_robust_waic <- because(
  equations = list(Y ~ X), data = robust_data,
  family = c(Y = student_t()), WAIC = TRUE, quiet = TRUE
)

because_compare(fit_gaussian_waic, fit_robust_waic)
```

With genuinely heavy-tailed data, the Student-*t* model will typically
have a lower WAIC than the Gaussian.

---

## Example 2: Beta Family for Proportions in (0, 1)

Continuous proportions — percentage vegetation cover, task completion
rates, dietary proportions — are bounded between 0 and 1. A Gaussian
model can predict values outside this range; the **Beta distribution**
is the natural choice.

The Beta distribution is parameterised here by its mean $\mu \in (0,1)$
and a precision (concentration) parameter $\phi > 0$:

$$Y \sim \text{Beta}(\mu \phi, (1-\mu)\phi)$$

With this parameterisation, the variance is $\mu(1-\mu)/(\phi + 1)$:
larger $\phi$ means less variance around $\mu$.

```{r beta_create}
beta_prop <- because_family(
  name = "beta_prop",

  # JAGS: dbeta(shape1, shape2)
  # shape1 = mu * phi, shape2 = (1 - mu) * phi
  jags_likelihood = paste0(
    "{response}[{i}] ~ dbeta(",
    "{mu}[{i}] * phi_{response}, ",
    "(1 - {mu}[{i}]) * phi_{response})"
  ),

  link = "logit",   # Logit link maps (-Inf, Inf) to (0, 1)

  # Prior on the precision (concentration) parameter
  extra_priors = c("phi_{response} ~ dgamma(1, 0.1)"),

  description = "Beta distribution for proportions in (0, 1) with logit link"
)
```

```{r beta_use}
set.seed(7)
N <- 100

# Predictor: grazing intensity (standardised)
Grazing <- rnorm(N, 0, 1)

# Response: vegetation cover proportion in (0, 1)
# True relationship on the logit scale
logit_mu <- -0.5 - 0.8 * Grazing
mu_true  <- 1 / (1 + exp(-logit_mu))
Cover    <- rbeta(N, shape1 = mu_true * 5, shape2 = (1 - mu_true) * 5)

prop_data <- data.frame(Cover = Cover, Grazing = Grazing)

fit_beta <- because(
  equations = list(Cover ~ Grazing),
  data      = prop_data,
  family    = c(Cover = beta_prop()),
  n.iter    = 15000,
  n.burnin  = 3000
)

summary(fit_beta)
# beta_Cover_Grazing: effect on the logit scale (negative = grazing reduces cover)
# phi_Cover: estimated precision (higher = lower variance around predicted mean)
```

```{r beta_me}
# Convert from logit scale to probability scale using marginal effects
me_beta <- marginal_effects(fit_beta)
print(me_beta)
# Effect: expected change in Cover (proportion units) per 1-unit increase in Grazing
```

---

## Example 3: Truncated Gaussian (Positive-Only Data)

Some biological quantities are strictly positive: body mass, flight
speed, enzyme concentration. A standard Gaussian model can assign
positive probability to negative values. JAGS supports truncation
notation `T(lower, upper)`.

```{r truncated_gaussian}
pos_gaussian <- because_family(
  name = "pos_gaussian",
  jags_likelihood = "{response}[{i}] ~ dnorm({mu}[{i}], {tau}) T(0, )",
  link = "log",   # Log link ensures the linear predictor maps to positive means
  description     = "Truncated Gaussian (positive support only)"
)

set.seed(99)
N <- 100
Mass <- rnorm(N, mean = 50, sd = 10)
Speed <- exp(1.2 + 0.05 * Mass + rnorm(N, 0, 0.3))  # Strictly positive

pos_data <- data.frame(Speed = Speed, Mass = scale(Mass)[, 1])

fit_pos <- because(
  equations = list(Speed ~ Mass),
  data      = pos_data,
  family    = c(Speed = pos_gaussian()),
  quiet     = TRUE
)

summary(fit_pos)
```

---

## Checking the Generated JAGS Code

You can inspect the JAGS model code that `because` generates for any
family to verify the substitution is correct before running a full
model:

```{r check_code}
model_code <- because_model(
  equations = list(Cover ~ Grazing),
  family    = c(Cover = beta_prop())
)

cat(model_code$model)
```

Look for the likelihood loop and confirm that `{response}`, `{mu}`, and
`{tau}` were correctly substituted.

---

## Advanced: The S3 Hook System for Extension Packages

`because_family()` is built on a general S3 hook system that allows
developers of extension packages to add full support for new families or
covariance structures.

The key generics are documented in `R/generics.R`:

| Generic | Purpose |
|:---|:---|
| `jags_family_definition` | Defines the JAGS likelihood block for a response variable |
| `jags_family_likelihood` | Generates the per-observation likelihood code (used by `because_family()`) |
| `jags_family_precision_prior` | Sets the prior on the precision/dispersion parameter |
| `prepare_structure_data` | Prepares custom covariance structures (e.g., trees, distance matrices) for JAGS |
| `get_inits_hook` | Provides specialised MCMC starting values |
| `normalize_equations_hook` | Handles specialised variable aliases |

### Registering a Family from an Extension Package

If you are building an extension package (e.g., `because.mytrait`), you
can register a custom family method using standard S3 registration:

```{r extension_registration}
# In your package's R/ directory, define the S3 method:
jags_family_likelihood.because_family_my_dist <- function(
    family, response, predictors = NULL, suffix = "", ...
) {
  list(
    likelihood_code = paste0(
      "    ", response, "[i] ~ dmy_dist(mu_", response, "[i], tau_res_", response, ")"
    ),
    prior_code = paste0(
      "  tau_res_", response, " ~ dgamma(1, 1)"
    ),
    data_requirements = NULL
  )
}

# Then register it so S3 dispatch works:
registerS3method(
  "jags_family_likelihood",
  "because_family_my_dist",
  jags_family_likelihood.because_family_my_dist,
  envir = asNamespace("because")
)
```

For custom **covariance structures** (e.g., spatial distance matrices),
use `because_structure()` — a parallel helper function described in
`?because_structure`.

---

## Summary

| Scenario | Tool | Key argument |
|:---|:---|:---|
| One-off custom distribution | `because_family()` | `jags_likelihood`, `extra_priors` |
| Heavy-tailed robust regression | `because_family()` with `dt()` | `df_{response}` extra prior |
| Proportions in (0,1) | `because_family()` with `dbeta()` | logit link, `phi_{response}` |
| Strictly positive data | `because_family()` with `T(0,)` truncation | log link |
| Custom covariance structure | `because_structure()` | `precision_fn` |
| Reusable family for a package | S3 method for `jags_family_likelihood` | Register with `registerS3method()` |

