---
title: "Counterfactual Simulations with the do() Operator"
author: "Achaz von Hardenberg"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Counterfactual Simulations with the do() Operator}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

## Introduction to Counterfactuals

One of the primary goals of causal inference is to answer "what-if" questions: *What would happen to the population size if we increased the temperature by 2°C? What if we reduced poaching by 50%?* 

In Pearl’s causal framework, these questions are formalized using the **$do()$-calculus**. When you apply an intervention to a variable (e.g., $do(X = x)$), you are essentially forcing that variable to take a specific value, "severing" it from its natural causes. The effect of this intervention then ripples downstream through the causal graph.

The `because` package natively supports Bayesian counterfactual simulations via the `do()` operator. Because the engine understands the structural equations and the topology of your DAG, it automatically propagates interventions correctly while carrying forward the full posterior uncertainty of the model.

```{r setup}
library(because)
```

## Simulating a Causal System

Let's set up a classic ecological cascade: **Temperature** drives the availability of a **Resource**, which in turn determines the **Abundance** of a species. We will also include a direct physiological effect of Temperature on Abundance.

```{r sim-data}
set.seed(42)
N <- 100

# True parameters
# Temperature (mean = 20C, sd = 4C)
raw_Temp <- rnorm(N, mean = 20, sd = 4)

# We scale variables for model fitting (Standard practice)
df <- data.frame(Site = 1:N)
df$Temp <- scale(raw_Temp) 

# Resource is driven by Temperature
df$Resource <- 0.6 * df$Temp + rnorm(N, 0, 1)

# Abundance is driven by both (Poisson distributed)
log_lambda <- 1.2 * df$Resource - 0.5 * df$Temp
df$Abundance <- rpois(N, lambda = exp(log_lambda))
```

We define our equations and fit the structural model:

```{r fit-model}
eqs <- list(
  Resource ~ Temp,
  Abundance ~ Resource + Temp
)

# Fit the model using JAGS (short chains for vignette speed)
fit <- because(eqs, data = df, 
               family = list(Abundance = "poisson"),
               n.iter = 500, n.burnin = 100, n.chains = 2, 
               quiet = TRUE)
```

## The `do()` Operator

The `do()` function takes a fitted `because` model and a set of targeted interventions. It returns a `because_counterfactual` object containing matrices of simulated posterior draws for every variable in the graph.

### 1. Atomic (Hard) Interventions
An atomic intervention forces a variable to a specific, fixed value across all observations. Because we scaled our data, `Temp = 2` means we are setting the temperature to exactly +2 standard deviations above the historical mean for every single site.

```{r hard-interv}
# Force Temperature to +2 SD at all sites
res_atomic <- do(fit, Temp = 2)

# Summarize the counterfactual predictions globally
summary(res_atomic)
```

Notice that the `SD` for `Temp` is exactly 0—because we intervened and forced it to be deterministically fixed. However, `Resource` and `Abundance` maintain their Bayesian uncertainty.

### 2. Shift (Additive) Interventions
Often, we don't want to force all sites to be identical. Instead, we want to shift the historical baseline. We can do this using a formula, where `.` represents the historical/natural values of the variable.

```{r shift-interv}
# Increase historical temperature by 1 standard deviation at every site
res_shift <- do(fit, Temp = ~ . + 1)
summary(res_shift)
```

## Intervening on the Raw Metric (`raw_scale = TRUE`)

There is a major conceptual trap when applying interventions to z-scored (scaled) data. As we just saw, adding `+1` to scaled data means adding 1 standard deviation, not 1 real-world unit.

If you assigned your scaled data safely (`df$Temp <- scale(raw_Temp)`), `because` secretly saved the original raw mean and standard deviation. By using `raw_scale = TRUE`, you can tell the `do()` operator to temporarily unscale the data, apply your intervention on the **raw biological metric** (like +2°C), and then re-scale it back before running the simulation!

```{r raw-scale}
# Increase the RAW temperature by exactly 2 degrees Celsius
# (The model handles the standard deviation math for you)
res_raw <- do(fit, Temp = ~ . + 2, raw_scale = TRUE)

summary(res_raw)
```

> **Important Note:** Base R's `data.frame(Temp = scale(raw_temp))` silently strips scaling attributes. To use `raw_scale = TRUE`, you must add scaled columns to existing dataframes using the `$` operator: `df$Temp <- scale(raw_temp)`.

### 3. Percentage Shifts
You can also apply multiplicative shifts easily by passing a character string ending in `%`. 

**However, remember our warning about scaled data!** If a variable has a mean of 0, multiplying it by 1.10 just stretches the variance. To safely apply a percentage shift, you should use the `raw_scale = TRUE` argument. This tells the engine to unscale the data to its natural positive values, apply the 10% increase, and automatically rescale it back before simulation.

```{r perc-interv}
# Increase the RAW historical temperature by 10% 
# (e.g., if it was 20C, it becomes 22C)
res_perc <- do(fit, Temp = "+10%", raw_scale = TRUE)
summary(res_perc)
```

*(Note: Under the hood, the `%` string syntax is a convenient shortcut that intercepts the unscaled matrix and multiplies it by `1 + (val/100)`).*

### 4. Stochastic Interventions
Sometimes policies are not perfectly exact. You can introduce noise into your intervention using standard R random number generators. Use `n` in the formula to represent the number of data points.

```{r stoch-interv}
# Shift temperature by +1 SD, but add observation noise to the intervention
res_stoch <- do(fit, Temp = ~ rnorm(n, mean = . + 1, sd = 0.2))
summary(res_stoch)
```

## Extracting Site-Specific Counterfactuals

The `summary()` function is great for looking at the **global expected value** (averaged across all sites). But often, you want to map or analyze how specific sites responded to the intervention.

The `do()` operator returns the full `[ndraws x N_obs]` matrices.

```{r site-specific}
# Extract the posterior matrix for Abundance under the raw shift
abund_matrix <- res_raw$Abundance

# Look at the dimensions: [posterior draws, sites]
dim(abund_matrix)

# Calculate the mean counterfactual abundance for Site 5
mean(abund_matrix[, 5])

# Calculate the 95% Credible Interval for Site 5
quantile(abund_matrix[, 5], probs = c(0.025, 0.975))
```

This flexibility allows you to easily compute site-specific treatment effects (the difference between the historical values and the counterfactual values) for precision ecology!
