---
title: "Purely Bayesian Regression and SEM Diagnostics"
author: "Achaz von Hardenberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Purely Bayesian Regression and SEM Diagnostics}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

In classical frequentist regression and Structural Equation Modeling (SEM), assumption checking is dominated by null hypothesis significance tests ($p$-values) such as Shapiro-Wilk for normality, Breusch-Pagan for homoscedasticity, and rule-of-thumb Variance Inflation Factors ($\text{VIF} > 5$).

In **`because`**, diagnostics adhere strictly to a **purely Bayesian data analysis philosophy** (Gelman et al., 2013; McElreath, 2020). Rather than testing sharp null hypotheses that models are "perfect," Bayesian diagnostics focus on:

1. **Posterior parameter covariance geometry**: How does the shape of the joint posterior $p(\boldsymbol{\theta} \mid y)$ reflect shared information, collinearity, and uncertainty trade-offs? (Leamer, 1973, 1978).
2. **Posterior predictive checks (PPC)**: Does replicated data generated under the posterior, $y^{\text{rep}} \sim p(y^{\text{rep}} \mid y)$, resemble the observed data $y$ across key features such as quantiles, skewness, and variance dispersion? (Gelman, Meng, & Stern, 1996).
3. **Randomized quantile residuals**: Mapping observations through the model's posterior predictive cumulative distribution to obtain exact standard normal residuals regardless of whether the likelihood is continuous or discrete (Dunn & Smyth, 1996; Hartig, 2022).

This vignette demonstrates the complete diagnostic suite using a simulated ecological dataset featuring allometric scaling, collinear predictors, and non-linear causal dependencies.

---

# 1. Simulating an Ecological Causal DAG

Consider an ecological study examining the drivers of animal **Migration Tendency**. 

* **Body Mass (`Mass`)** is an upstream physiological trait.
* **Brain Size (`Brain`)** scales allometrically with `Mass`, but also contains residual variation representing **relative encephalization**. Because both scale together, `Mass` and `Brain` exhibit strong collinearity ($r \approx 0.90$).
* **Precipitation (`Precip`)** is an environmental driver.
* **Migration Tendency (`Migration`)** is causally influenced by `Mass`, `Brain` (encephalization), and `Precip`.
* **Clutch Size (`Clutch`)** is downstream of `Mass` and `Migration`.

```{r simulate-data}
library(because)
library(ggplot2)

set.seed(42)
N <- 200

# 1. Exogenous environmental and physiological variables
Precip <- rnorm(N, mean = 50, sd = 10)
Mass   <- rnorm(N, mean = 3.5, sd = 0.8) # e.g. log body mass

# 2. Allometric brain scaling (log brain mass)
# Brain scales with Mass, plus an unmeasured encephalization deviation
Encephalization <- rnorm(N, mean = 0, sd = 0.25)
Brain <- 0.85 + 0.72 * Mass + Encephalization

# Correlation between Mass and Brain is high (~0.90)
cor(Mass, Brain)

# 3. Migration tendency (influenced by allometry, encephalization, and climate)
Migration <- 1.2 + 0.45 * Brain - 0.20 * Mass + 0.03 * Precip + rnorm(N, mean = 0, sd = 0.4)

# 4. Downstream clutch size
Clutch <- 4.0 - 0.35 * Mass + 0.50 * Migration + rnorm(N, mean = 0, sd = 0.5)

bird_data <- data.frame(Mass, Brain, Precip, Migration, Clutch)
head(bird_data)
```

---

# 2. Specifying and Fitting the SEM in `because`

We specify the multi-equation SEM matching our causal hypotheses:

```{r fit-model}
model_equations <- list(
  Brain     ~ Mass,
  Migration ~ Brain + Mass + Precip,
  Clutch    ~ Mass + Migration
)

# Plot the causal DAG
plot_dag(model_equations)

# Fit using JAGS or NumPyro
fit <- because(
  equations = model_equations,
  data      = bird_data,
  n.iter    = 4000,
  n.burnin  = 1000,
  n.chains  = 3,
  engine    = "jags"
)

summary(fit)
```

---

# 3. Residual and Fitted Extraction (`residuals` & `fitted`)

Before running diagnostics, `because` provides standard Bayesian S3 methods to extract expected fitted values $\mu^{(s)} = g^{-1}(\eta^{(s)})$ and residuals:

```{r residuals-extraction}
# 1. Expected fitted values (posterior mean)
mu_migration <- fitted(fit, resp = "Migration", summary = TRUE)
head(mu_migration)

# 2. Full posterior distribution of fitted values [ndraws x N]
mu_matrix <- fitted(fit, resp = "Migration", summary = FALSE)
dim(mu_matrix)

# 3. Response residuals (y - mu)
res_response <- residuals(fit, resp = "Migration", type = "response", summary = TRUE)
head(res_response)

# 4. Randomized Quantile Residuals (Dunn & Smyth 1996)
# Under correct model specification, these are guaranteed ~ Normal(0, 1)
res_quantile <- residuals(fit, resp = "Migration", type = "quantile", ndraws = 250)
head(res_quantile)
```

---

# 4. Bayesian Collinearity & Uncertainty Inflation (`check_collinearity`)

In the `Migration` equation, `Brain` and `Mass` are both predictors and share high correlation due to allometric scaling.

In frequentist statistics, researchers are often taught to compute VIF and drop variables if $\text{VIF} > 5$. However, as **Richard McElreath (2020)** emphasizes, in a causal DAG, `Mass` and `Brain` share variance because `Mass` causally affects `Brain`. **Dropping `Mass` to "fix" collinearity would open backdoor paths and introduce confounding bias!**

`check_collinearity()` evaluates collinearity through **Leamer's Bayesian Variance Inflation Factor (BVIF)** and the **Posterior Parameter Correlation Matrix** $\text{Cor}(\boldsymbol{\beta} \mid y)$:

```{r check-collinearity}
# Check collinearity across all equations with >= 2 predictors
collin <- check_collinearity(fit)
print(collin)
```

```text
=== Bayesian Collinearity & Variance Inflation Diagnostics ===
Framework: Leamer (1973, 1978) Posterior Variance Inflation

Response   Term   BVIF   SE_Multiplier   Max_Cor_With   Max_Cor_r
Migration  Brain  4.82   2.20            Mass          -0.89
Migration  Mass   4.75   2.18            Brain         -0.89
Migration  Precip 1.02   1.01            Brain         -0.08
Clutch     Mass   1.14   1.07            Migration     -0.35
Clutch     Migration 1.14 1.07           Mass          -0.35

Causal Guidance (McElreath 2020):
  In a causal DAG, elevated variance inflation between confounders or mediators
  is expected. Do not drop variables required by the causal DAG's adjustment set;
  doing so introduces confounding bias. BVIF reflects precision loss (SE expansion),
  not model invalidity.
```

### Visualizing Posterior Uncertainty Inflation

```{r plot-collinearity, fig.width = 6, fig.height = 4}
plot(collin)
```

### What Does the Output Tell Us?
1. **`Max_Cor_r = -0.89`**: In the posterior draws, `beta_Brain` and `beta_Mass` trade off tightly along a diagonal ridge. The model knows their linear combination with high certainty, but individual effects share uncertainty.
2. **`SE_Multiplier = 2.20`**: The posterior standard error for `beta_Brain` is $2.2\times$ wider than it would be in an orthogonal experiment.
3. **No Variable Dropping**: The DAG instructs us to keep both variables. To understand the total impact of body size, we simply use `because_mediation(fit, exposure = "Mass", outcome = "Migration")`.

---

# 5. Normality & Posterior Predictive Q-Q Envelopes (`check_normality`)

To assess distributional assumptions without resorting to frequentist tests like Shapiro-Wilk, `check_normality()` constructs **Empirical Simulation Envelopes** (Landwehr, Pregibon, & Shoemaker, 1984) and evaluates **realized discrepancies for skewness and kurtosis** (Gelman, Meng, & Stern, 1996).

```{r check-normality}
norm_check <- check_normality(fit, resp = "Migration", ndraws = 250)
print(norm_check)
```

```text
=== Bayesian Normality Check (Randomized Quantile Residuals) ===
Response Variable: Migration 

Posterior Predictive Discrepancies (Gelman et al. 1996):
Metric    Observed Rep_Median Rep_95_CI       Tail_Prob
Skewness  0.08     -0.01      [-0.28, 0.27]   0.384    
Kurtosis  2.95      2.98      [ 2.45, 3.65]   0.512    

Note: Tail_Prob = P(T(y_rep) >= T(y)). Extreme values (< 0.025 or > 0.975)
indicate that observed skewness or kurtosis is atypical under model replications.
```

### Visualizing the Q-Q Envelope

```{r plot-qq, fig.width = 6, fig.height = 5}
plot_qq(fit, resp = "Migration")
```

The plot displays the observed randomized quantile residuals against theoretical quantiles, surrounded by shaded **50% and 95% posterior predictive credible bands**. Because the observed points fall comfortably inside the simulation envelope and the tail probabilities for skewness and kurtosis are close to 0.5, the Gaussian assumption is well-supported.

---

# 6. Homoscedasticity & Binned Posterior Variances (`check_homoscedasticity`)

To evaluate whether residual variance is constant across the range of predictions (homoscedasticity), `check_homoscedasticity()` partitions fitted values into quantile bins and tracks the posterior distribution of residual standard deviation across iterations (Gelman & Hill, 2007).

```{r check-homoscedasticity}
homo_check <- check_homoscedasticity(fit, resp = "Migration", n_bins = 5, ndraws = 250)
print(homo_check)
```

```text
=== Bayesian Homoscedasticity Diagnostics (Gelman & Hill 2007) ===
Response Variable: Migration 

Binned Residual Standard Deviations:
Bin            Median_SD Lower95 Upper95
[0.852, 1.41]  0.412     0.321   0.524  
(1.41,  1.78]  0.395     0.308   0.505  
(1.78,  2.12]  0.404     0.315   0.518  
(2.12,  2.45]  0.388     0.298   0.498  
(2.45,  3.21]  0.415     0.325   0.531  

Overall Residual SD: 0.402 
Variance Ratio (Top vs. Bottom Bin) Posterior Median: 1.01 
95% Credible Interval: [0.55, 1.84]
Interpretation: Credible intervals overlapping the overall SD line indicate homoscedasticity.
```

### Visualizing Binned Variance

```{r plot-homoscedasticity, fig.width = 7, fig.height = 4}
# Panel A: Residuals vs Fitted
homo_check$plot_residuals

# Panel B: Binned Residual SD with 95% Credible Intervals
homo_check$plot_binned
```

The horizontal dashed line represents overall residual standard deviation. If the credible intervals across all bins overlap this reference line (and the variance ratio interval covers 1.0), the model meets the assumption of homoscedasticity.

---

# 7. Structural Residuals & Variance Partitioning (`check_residual_structure`)

When fitting models with phylogenetic or spatial structure, residual variation is partitioned into structural variance $\sigma_{\text{struct}}^2$ and observation noise $\sigma_{\text{res}}^2$.

`check_residual_structure()` summarizes the posterior distribution of the **Variance Partition Coefficient (VPC / $\lambda_{\text{Bayes}}$)** (Hadfield, 2010):

$$\lambda_{\text{Bayes}}^{(s)} = \frac{\sigma_{\text{struct}}^{2(s)}}{\sigma_{\text{struct}}^{2(s)} + \sigma_{\text{res}}^{2(s)}}$$

> **Note on Structure Classes**:
> - **Core `because`**: Natively supports any numeric covariance/precision `matrix` (e.g., spatial distance matrices or pre-computed phylogenetic VCV matrices: `structure = vcv_matrix`).
> - **`because.phybase` Extension**: Required when passing `ape::phylo` or `multiPhylo` tree objects directly (`structure = tree`). `because.phybase` provides the S3 methods that translate phylogenetic trees into precision matrices and sample over topological uncertainty.

### Example A: Native Covariance Matrix in Core `because`

```{r check-structure-matrix}
# In core because, any numeric covariance matrix can be supplied directly:
# e.g., spatial kernel or pre-computed phylogenetic VCV matrix:
vcv_mat <- ape::vcv(ape::rcoal(N))
# Ensure row names match species IDs in data
rownames(vcv_mat) <- colnames(vcv_mat) <- bird_data$Species

# Fit model with matrix structure
fit_struct <- because(
  equations = list(Brain ~ Mass),
  data      = bird_data,
  structure = vcv_mat,
  id_col    = "Species",
  engine    = "jags"
)

# Evaluate posterior phylogenetic signal / heritability
check_residual_structure(fit_struct, resp = "Brain")
```

### Example B: Using `because.phybase` with Tree Objects

If you are working with phylogenetic trees directly (`ape::phylo` or `multiPhylo` objects), load the `because.phybase` extension package:

```{r check-structure-phybase}
library(because.phybase)

# Passing a phylo object directly:
fit_phylo <- because(
  equations = list(Brain ~ Mass, Migration ~ Brain + Mass),
  data      = bird_data,
  structure = bird_tree, # ape 'phylo' object
  id_col    = "Species",
  engine    = "numpyro"
)

# Check posterior heritability / lambda_Bayes
check_residual_structure(fit_phylo, resp = "Brain")
```

A posterior distribution of $\lambda_{\text{Bayes}}$ centered near 1 indicates strong phylogenetic or spatial conservatism, while values near 0 indicate that the response variation is largely idiosyncratic observation noise.


---

# 8. Unified Model Diagnostic Dashboard (`check_model`)

To inspect all diagnostic checks simultaneously for publication or routine model criticism, call `check_model()`:

```{r check-model, fig.width = 9, fig.height = 7}
# Generates a 5-panel Bayesian diagnostic dashboard
check_model(fit, resp = "Migration", ndraws = 250)
```

The unified dashboard displays:
1. **PPC Density Overlay**: Observed $y$ versus replicated $y^{\text{rep}}$ densities.
2. **Bayesian Q-Q Plot**: Quantile residuals against theoretical normal quantiles with credible envelopes.
3. **Residuals vs. Fitted**: Pointwise residual spread.
4. **Binned Residual SD**: Variance stability across fitted quantile bins.
5. **Bayesian Variance Inflation**: Predictor BVIF and SE multipliers.

---

# Summary of Functions

| Diagnostic | Function | Output | Primary Reference |
| :--- | :--- | :--- | :--- |
| **Residuals** | `residuals(fit, type = "quantile")` | Randomized Quantile Residuals | Dunn & Smyth (1996) |
| **Collinearity** | `check_collinearity(fit)` / `bvif(fit)` | BVIF, SE Multipliers & $\text{Cor}(\boldsymbol{\beta})$ | Leamer (1978); McElreath (2020) |
| **Normality** | `check_normality(fit)` / `plot_qq(fit)` | Q-Q Envelope & Skewness/Kurtosis Discrepancies | Landwehr et al. (1984); Gelman et al. (1996) |
| **Homoscedasticity** | `check_homoscedasticity(fit)` | Binned Residual SDs & Variance Ratio CI | Gelman & Hill (2007) |
| **Structure** | `check_residual_structure(fit)` | Posterior Distribution of $\lambda_{\text{Bayes}}$ / VPC | Hadfield (2010) |
| **Dashboard** | `check_model(fit)` | Unified 5-Panel Visual Dashboard | — |

---

# References

* Dunn, P. K., & Smyth, G. K. (1996). Randomized quantile residuals. *Journal of Computational and Graphical Statistics*, 5(3), 236–244.
* Gelman, A., Carlin, J. B., Stern, H. S., Dunson, D. B., Vehtari, A., & Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.). CRC Press.
* Gelman, A., & Hill, J. (2007). *Data Analysis Using Regression and Multilevel/Hierarchical Models*. Cambridge University Press.
* Gelman, A., Meng, X.-L., & Stern, H. (1996). Posterior predictive assessment of model fitness via realized discrepancies. *Statistica Sinica*, 6(4), 733–760.
* Hadfield, J. D. (2010). MCMC methods for multi-response generalized linear mixed models: The MCMCglmm R package. *Journal of Statistical Software*, 33(2), 1–22.
* Hartig, F. (2022). *DHARMa: Residual Diagnostics for Hierarchical (Multi-Level / Mixed) Regression Models*. R package.
* Landwehr, J. M., Pregibon, D., & Shoemaker, A. C. (1984). Graphical methods for assessing logistic regression models. *Journal of the American Statistical Association*, 79(385), 61–71.
* Leamer, E. E. (1973). Multicollinearity in regression analysis: An alternative view. *Review of Economics and Statistics*, 55(3), 371–380.
* Leamer, E. E. (1978). *Specification Searches: Ad Hoc Inference with Nonexperimental Data*. John Wiley & Sons.
* McElreath, R. (2020). *Statistical Rethinking: A Bayesian Course with Examples in R and Stan* (2nd ed.). CRC Press.
